19Q2

### 2019-05-15 19Q2.Rmd

---
title: "19Q1"
author: "Sally Claridge"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    theme: cerulean
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
    code_folding: hide
    pandoc_args: [
      "+RTS", "-K2048m",
      "-RTS"
    ]
---

@@

Note that the new DepMap 19Q1 release made multiple changes to variable names:

- For their universal ccancer cell line nomenclature, they reverted back to `Broad_ID` from 18Q3 rather than `DepMap_ID` that was used in 19Q1
- The `Primary.Disease` column of the cell line metadata file is more consistently named (`Subtype.Disease` now has the specific disease annotations), though there are a couple capitalization inconsistencies
- MAF file no longer has `CCLE_Name` column and has a duplicate column of the Broad IDs (i.e. `DepMap_ID` and `Tumor_Sample_Barcode`)

And changes to the data processing pipelines (the following is taken verbatim from the [DepMap data portal](https://depmap.org/portal/download/)):

- **Expression**: We use the GTEx pipeline. Since last release, there have been some updates to the versions of different components of the pipeline and minor bug fixes on GTEx's end.
- **Mutation**: In this release, we additionally process DepMap WES samples through the CGA mutation calling pipeline adapted for cell lines and available on FireCloud. We have added a column for the allele counts from this pipeline run (CGA_WES_AC). Previously, the WES_AC column was generated from running the Van Allen mutation calling pipeline for SNPs + Mutect2 for INDELs. This column is now named VA_WES_AC.
- **Copy Number**: We have moved to using the newest version of the GATK pipeline.

@@

This comprehensive cancer cell line information was curated by both Daniel Charytonowicz and myself.
```{r dan_ccl_database}
ccl_converter <- read.delim("./data_munging/cell_line_database_v1_20180911.tsv", row.names = 1, sep = "\t", header = TRUE)
ccl_lineage_converter <- unique(ccl_converter[, c("Broad_ID", "grouped_general_doid", "group_general_lineage_name")])
lineage_converter <- unique(select(ccl_lineage_converter, grouped_general_doid, group_general_lineage_name))
ccl_histologies <- read.delim("./data_munging/ccl_histologies_metadata.csv", sep = ",", header = TRUE, row.names = 1)
```

@@

DepMap 19Q1 update fixed almost all of the inconsistencies of the Primary.Disease column, though there were come capitalization inconsistencies.
```{r depmap_metadata}
ccl_meta <- read.delim("./data_munging/19Q1/DepMap-2019q1-celllines_v2_20190305.csv", sep = ",", header = TRUE, na.strings = c("", NA))
ccl_meta$Primary.Disease <- tolower(ccl_meta$Primary.Disease)
ccl_meta$Primary.Disease <- with(ccl_meta, ifelse(Primary.Disease == "colon/colorectal cancer", "colorectal cancer", Primary.Disease))
ccl_meta <- subset(ccl_meta, Primary.Disease != "unknown" & Primary.Disease != "primary cells" & !grepl("MERGED", CCLE_Name, fixed = TRUE))
ccl_meta$Cell_line <- as.character(lapply(ccl_meta$CCLE_Name, function(x) strsplit(as.character(x), split = "_")[[1]][1]))
ccl_meta$Histology <- tolower(gsub("^[a-z0-9]*_", "", ccl_meta$CCLE_Name, ignore.case = TRUE))
# Convert Primary.Disease entries to all lower case to fix factorization issue
ccl_meta$Primary.Disease <- as.factor(tolower(ccl_meta$Primary.Disease))
# Harmonize ID column name
colnames(ccl_meta)[colnames(ccl_meta) == "DepMap_ID"] <- "Broad_ID"

# Remove two rows that contain cell lines with two different histologies
# The histologies that are kept are verified through matching in Cellosaurus and Cell Model Passports
# KMH2, haematopoietic_and_lymphoid: https://web.expasy.org/cellosaurus/CVCL_1330, https://cellmodelpassports.sanger.ac.uk/passports/SIDM01018
# MS1, lung: https://web.expasy.org/cellosaurus/CVCL_1429, https://cellmodelpassports.sanger.ac.uk/passports/SIDM00245
ccl_meta <- subset(ccl_meta, CCLE_Name != "KMH2_THYROID" & CCLE_Name != "MS1_SKIN")
```

@@

```{r g2p_parsing_functions}
ExtractBiomarkerType <- function(json_string) {
  json_string <- as.character(json_string)
  biomarker_type <- ((str_extract(string = json_string, pattern = "biomarker_type.+") %>% strsplit(split = ","))[[1]][1] %>% str_extract(pattern = "u'.+") %>% str_match("'(.*?)'"))[,2]
  return(biomarker_type)
}
```

```{r g2p_parsing, eval=FALSE}
# g2p_json <- "./../htscreens_giant_files/g2p_all.json"
# g2p_all <- fromJSON(sprintf("[%s]", paste(readLines(g2p_json), collapse = ",")), flatten = TRUE)
# saveRDS(g2p_all, "./../htscreens_giant_files/g2p_all_flattened_json.rds", compress = "xz")

g2p <- read.delim("./data_munging/g2p_full_level_A_dataset_NEW_DEC_2_2018_withGroupedLineages.csv", sep = "\t", header = TRUE)
g2p[, 1:2] <- NULL
g2p <- unfactorize(g2p)
g2p$Drug <- tolower(g2p$Drug)
g2p[g2p == -1] <- NA
g2p$Gene <- ifelse(is.na(g2p$Gene) & g2p$Drug == "daunorubicin", "MLL", as.character(g2p$Gene))
g2p$Direction <- ifelse(g2p$Direction == 1 | g2p$Direction == "", NA, as.character(g2p$Direction))
g2p$Ref[g2p$Ref == "None" | g2p$Ref == "-" | is.na(g2p$Ref)] <- ""
g2p$Alt[g2p$Alt == "None" | g2p$Alt == "-" | is.na(g2p$Alt)] <- ""
g2p <- g2p[, c("Drug", "DrugID", "Direction", "Gene", "Ref", "Alt", "Start", "End", "Chromosome", "Feature.Names", "Mutation", "JSON", "Evidence.Level", "Phenotype.Description", "Phenotype.Family", "Phenotype.ID", "Sequence.ID", "grouped_general_doid")]
g2p <- add_column(g2p, Ref_Length = nchar(as.character(g2p$Ref)), .after = "Ref")
g2p <- add_column(g2p, Alt_Length = nchar(as.character(g2p$Alt)), .after = "Alt")
g2p <- add_column(g2p, Variant_Length = NA, .after = "Alt_Length")
g2p$Variant_Length <-
  ifelse(g2p$Alt_Length == 1 & g2p$Ref_Length == 1, 1,
  ifelse(g2p$Alt_Length == 2 & g2p$Ref_Length == 2, 2,
  ifelse(g2p$Alt_Length == 3 & g2p$Ref_Length == 3, 3,
  ifelse(g2p$Ref_Length == g2p$Alt_Length & g2p$Ref_Length > 3, g2p$Ref_Length,
  ifelse(g2p$Ref_Length != g2p$Alt_Length, abs(g2p$Ref_Length - g2p$Alt_Length), NA)))))
g2p <- add_column(g2p, Variant_Type = NA, .after = "Alt_Length")
g2p$Variant_Type <-
  ifelse(g2p$Alt_Length == 1 & g2p$Ref_Length == 1, "SNP",
  ifelse(g2p$Alt_Length == 2 & g2p$Ref_Length == 2, "DNP",
  ifelse(g2p$Alt_Length == 3 & g2p$Ref_Length == 3, "TNP",
  ifelse(g2p$Ref_Length == g2p$Alt_Length & g2p$Ref_Length > 3, "ONP",
  ifelse(g2p$Ref_Length > g2p$Alt_Length, "DEL",
  ifelse(g2p$Ref_Length < g2p$Alt_Length, "INS", "None"))))))
g2p <- subset(g2p, grepl("^CID", g2p$DrugID))
g2p <- add_column(g2p, biomarker_type = tolower(sapply(g2p$JSON, ExtractBiomarkerType, simplify = TRUE)), .after = "Feature.Names")
g2p <- add_column(g2p, biomarker_type_curated = NA, .after = "biomarker_type")
# Manually assign biomarker_types based on a combination of biomarker_type derived from the JSON string and patterns in the Feature.Names
biomarker_type_curated <-
  ifelse(!is.na(str_match(string = g2p$Feature.Names, pattern = "splice")) | !is.na(str_match(string = g2p$biomarker_type, pattern = "splice")), "Splice_Site",
  ifelse(!is.na(str_match(string = g2p$Feature.Names, pattern = "Amplification|amplification")), "amplification",
  ifelse(!is.na(str_match(string = g2p$Feature.Names, pattern = "^PDGFRA [(]")) | !is.na(str_match(string = tolower(g2p$Feature.Names), pattern = "mutant|mutation|mut")), "mutation",
  ifelse(!is.na(str_match(string = tolower(g2p$Feature.Names), pattern = "fusion")) | !is.na(str_match(string = tolower(g2p$biomarker_type), pattern = "fusion")), "fusion", 
  ifelse(!is.na(str_match(string = g2p$Feature.Names, pattern = "[A-Z0-9|A-Z]+-[A-Z0-9|A-Z]+")) & is.na(str_match(string = g2p$Feature.Names, pattern = "[0-9]+-[0-9]+")), "fusion",
  ifelse(g2p$Variant_Type == "DEL" & g2p$Variant_Length %% 3 == 0, "In_Frame_Del",
  ifelse(g2p$Variant_Type == "DEL" & g2p$Variant_Length %% 3 != 0, "Frame_Shift_Del",
  ifelse(g2p$Variant_Type == "INS" & g2p$Variant_Length %% 3 == 0, "In_Frame_Ins",
  ifelse(g2p$Variant_Type == "INS" & g2p$Variant_Length %% 3 != 0, "Frame_Shift_Ins",
  ifelse(!is.na(str_match(string = g2p$Feature.Names, pattern = "EGFR E746_A750del")), "In_Frame_Del",
  ifelse(!is.na(str_match(string = g2p$Feature.Names, pattern = "EGFR H773>PNPY")), "In_Frame_Ins",
  ifelse(g2p$Feature.Names == "KRAS G12G" | g2p$Feature.Names == "KRAS G13G", "Silent",
  ifelse(!is.na(str_match(string = g2p$Feature.Names, pattern = "[A-Z][0-9]+[A-WYZ]")) | !is.na(str_match(string = g2p$biomarker_type, pattern = "missense")), "Missense_Mutation",
  ifelse(!is.na(str_match(string = g2p$Feature.Names, pattern = "[A-Z][0-9]+([X]|[*])")), "Nonsense_Mutation",
  ifelse(!is.na(str_match(string = g2p$Feature.Names, pattern = "positive")), "positive",
  ifelse(!is.na(str_match(string = g2p$Feature.Names, pattern = "expression")), "expression",
  ifelse(g2p$biomarker_type == "rearrange", "rearrangement",
  ifelse(g2p$biomarker_type == "intron variant", "Intron",
  ifelse(g2p$biomarker_type == "inframe_deletion" | !is.na(str_match(string = tolower(g2p$Feature.Names), pattern = "inframe deletion")), "In_Frame_Del",
  ifelse(g2p$biomarker_type == "inframe_insertion" | !is.na(str_match(string = tolower(g2p$Feature.Names), pattern = "inframe insertion")), "In_Frame_Ins",
  ifelse(!is.na(str_match(string = tolower(g2p$Feature.Names), pattern = "biallelic inactivation")), "biallelic inactivation", NA)))))))))))))))))))))
g2p$biomarker_type_curated <- biomarker_type_curated[,1]
g2p$Drug_Gene <- ifelse(is.na(g2p$Drug) | is.na(g2p$Gene), NA, paste0(g2p$Drug, "_", g2p$Gene))
g2p$position <- paste0("g.chr", g2p$Chromosome, ":", g2p$End, g2p$Ref, ">", g2p$Alt)
g2p$position <- ifelse(grepl("NA", g2p$position), NA, g2p$position)
g2p <- merge(lineage_converter, g2p, by = "grouped_general_doid", all.y = TRUE)
g2p$Drug_Gene_Lin <- ifelse(is.na(g2p$Drug) | is.na(g2p$Gene) | is.na(g2p$group_general_lineage_name), NA, paste0(g2p$Drug, "_", g2p$Gene, "_", g2p$group_general_lineage_name))

# write.csv(g2p, "./data_munging/g2p_db_edited.csv")

# g2p_dgls <-  unique (g2p$Drug_Gene_Lin[!is.na(g2p$Drug_Gene_Lin)])
# g2p_druggenes <- unique(g2p$Drug_Gene[!is.na(g2p$Drug_Gene)])
# g2p_drugs <- unique(sapply(strsplit(g2p_dgls, "_", fixed = TRUE), function(x) x[1]))
# g2p_genes <- unique(sapply(strsplit(g2p_dgls, "_", fixed = TRUE), function(x) x[2]))
# g2p_lins <- unique(sapply(strsplit(g2p_dgls, "_", fixed = TRUE), function(x) x[3]))
# g2p_druggenes
# g2p_drugs
# g2p_genes
# g2p_lins

# Order and genes taken from pink heatmap from Daniel's analyses
# g2p_genes <- c("EGFR", "KRAS", "ERBB2", "ERBB3", "MET", "ALK", "ROS1", "PDGFRA", "ABL1", "KIT", "PDGFRB", "PDGFB", "BRAF", "G6PD", "RET", "BRCA1", "BRCA2", "UGT1A1", "TSC1", "TSC2", "AKT1", "IDH2", "DPYD", "ESR1", "CYP19A1", "TPMT", "FLT3", "NPM1", "DNMT3A", "PML", "PTCH1", "JAK2", "MYD88", "MGMT")
```

```{r g2p}
g2p <- read.delim("./data_munging/g2p_db_edited.csv", sep = ",", header = TRUE)

@@

```{r testing, eval=FALSE}
g2p$JSON[1]
test <- filter(g2p, is.na(Gene))
test$JSON[2]

unique(g2p$group_general_lineage_name)
levels(g2p$group_general_lineage_name)

g2p_nonCID <- subset(g2p, !grepl("^CID", g2p$DrugID))
summ <- g2p_nonCID %>% group_by(Drug, DrugID) %>% tally() %>% summarize(DrugID = DrugID, n = n, percent = formatC(n / 402 * 100, digits = 6, format = "f"))
summ
```

@@

Mutation calls (coding region, germline filtered):
```{r mutation_status, eval=FALSE}
maf_raw <- read.delim("./data_munging/19Q1/depmap_19Q1_mutation_calls_20190305.csv.gz", header = TRUE, sep = ",", row.names = 1)
# Harmonize ID column name
colnames(maf_raw)[colnames(maf_raw) == "DepMap_ID"] <- "Broad_ID"
colnames(maf_raw)[colnames(maf_raw) == "Hugo_Symbol"] <- "Gene"
# Select columns
maf_df <- subset(maf_raw, select = c("Gene", "Broad_ID", "Variant_Classification", "Reference_Allele", "Tumor_Seq_Allele1"))
maf_df$Reference_Allele <- ifelse(maf_df$Reference_Allele == "-", 0, nchar(as.character(maf_df$Reference_Allele)))
maf_df$Tumor_Seq_Allele1 <- nchar(as.character(maf_df$Tumor_Seq_Allele1))
maf_df <- maf_df[maf_df$Variant_Classification != "Silent",]
# Add Mutation_Status column
maf_df$Mutation_Status <- 1
maf_df$PointMutation <- with(maf_df, ifelse(Reference_Allele == 1 | is.na(Reference_Allele), TRUE, ifelse(Reference_Allele == 0 & Tumor_Seq_Allele1 == 1, 1, 0)))
maf_df <- unfactorize(maf_df)

saveRDS(maf_df, "./data_munging/19Q1/maf_df_19Q1.rds", compress = "xz")
```

@@

```{r fusion_calls_19Q2, eval=FALSE}
fusions <- read.delim("./data_munging/19Q2/CCLE_fusions.csv.gz", header = TRUE, sep = ",")
fusions$LeftGene <- gsub(" .*", "", fusions$LeftGene)
fusions$RightGene <- gsub(" .*", "", fusions$RightGene)
fusions$X.FusionName <- gsub("--", "-", fusions$X.FusionName)
fusions <- select(fusions, DepMap_ID, X.FusionName)
fusions <- as.data.frame(lapply(fusions, function(x) gsub("IGH@|IGH-@", "IGH", x)))
colnames(fusions) <- c("Broad_ID", "FusionPair")
```

DepMap WES CN Data:
```{r copy_number, eval=FALSE}
cn <- read.delim("./data_munging/19Q1/public_19Q1_gene_cn_20190305.csv.gz", sep = ",", check.names = FALSE, header = TRUE)
# Remove Entrez gene IDs from colnames, format is "Gene (Entrez ID)"
colnames(cn) <- gsub(" .*", "", colnames(cn))
colnames(cn)[1] <- "Broad_ID"
# Melt
cn_melt <- melt(data = cn, id.vars = "Broad_ID", measure.vars = colnames(cn[2:ncol(cn)]), variable.name = "Gene", value.name = "CN")
cn_melt <- unfactorize(cn_melt)

saveRDS(cn_melt, "./data_munging/19Q1/cn_melt_19Q1.rds", compress = "xz")
```

19Q1, CCLE RNAseq gene expression data:
```{r gene_expression, eval=FALSE}
ge <- read.delim("./../htscreens_giant_files/CCLE_depMap_19Q1_TPM_20190305.csv.gz", header = TRUE, sep = ",", check.names = FALSE)
# Remove Entrez gene IDs from colnames, format is "Gene (Entrez ID)"
colnames(ge) <- gsub(" .*", "", colnames(ge))
colnames(ge)[1] <- "Broad_ID"
# Melt
ge_melt <- melt(ge, id.vars = "Broad_ID", measure.vars = colnames(ge)[2:ncol(ge)], variable.name = "Gene", value.name = "TPM")
ge_melt$TPM <- log2(ge_melt$TPM + 1)
ge_melt <- unfactorize(ge_melt)

saveRDS(ge_melt, "./data_munging/19Q1/ge_melt_19Q1.rds", compress = "xz")
```

Load Rds:
```{r load_omics_rds}
maf_df <- readRDS("./data_munging/19Q1/maf_df_19Q1.rds")
cn_melt <- readRDS("./data_munging/19Q1/cn_melt_19Q1.rds")
ge_melt <- readRDS("./data_munging/19Q1/ge_melt_19Q1.rds")
```

```{r omics_testing, eval=FALSE}
test <- filter(maf_df, Gene %in% g2p_genes) %>% count(Gene, Broad_ID)
test <- test[order(test$n, decreasing = TRUE),]
```

@@

```{r omics_functions}
MergeOmicsG2P <- function(database, maf_df, ge_df, cn_df, outfile) {
  # Merge mutation status, CN, and GE omics data with drugscreen results and metadata
  if(database %in% c("CCLE", "CTRP", "GDSC")) {
    filt <- auc %>% filter(Database == database) %>% select(Cell_line, Histology, TCGA_classification, CCLE_Name, Broad_ID, Drug, AUC_IC50)
    print(paste0("Processing ", database, " data."))
    colnames(filt)[colnames(filt) == "AUC_IC50"] <- "Score"

    # Get all G2P gene-cell line combinations
    grid <- expand.grid("Cell_line" = unique(filt$Cell_line), "Gene" = g2p_genes)
    data <-  merge(filt, grid, by = "Cell_line", all.y = TRUE)
    print("Merged grid.")
    # Merge metadata
    data <-  merge(data, select(ccl_meta, Cell_line, Histology, CCLE_Name, Broad_ID, Primary.Disease), by = c("Cell_line", "Histology", "CCLE_Name", "Broad_ID"), all.y = TRUE)
    ## Add columns for G2P drug-gene and drug-gene-lineage sorting
    data$Drug_Gene <- paste0(data$Drug, "_", data$Gene)
    data$G2P_Drug_Gene <- ifelse(data$Drug_Gene %in% g2p_druggenes, "Yes", "No")
    data$Drug_Gene_Lin <- paste0(data$Drug, "_", data$Gene, "_", data$Primary.Disease)
    data$G2P_Drug_Gene_Lin <- ifelse(data$Drug_Gene_Lin %in% g2p_dgls, "Yes", "No")
    print("Merged metadata.")
    # Merge mutation calls
    data <- unfactorize(data)
    data <- merge(data, maf_df, by = c("Gene", "Broad_ID"), all.x = TRUE)
    data$Mutation_Status <- ifelse(is.na(data$Mutation_Status), 0, data$Mutation_Status)
  }
  if(database == "CRISPR") {
    print(paste0("Processing ", database, " data."))
    # Melt CRISPR dataset for merging
    data <- melt(crispr_all, id.vars = "Broad_ID", measure.vars = colnames(crispr_all)[2:ncol(crispr_all)], variable.name = "Gene", value.name = "Score")
    # Get all G2P gene-cell line combinations
    data <-  filter(data, Gene %in% g2p_genes)
    # Merge cell line metadata
    data <- merge(data, select(ccl_meta, Cell_line, Histology, CCLE_Name, Broad_ID, Primary.Disease), by = "Broad_ID", all.y = TRUE)
    data$Drug_Gene <- paste0(data$Drug, "_", data$Gene)
    data$Drug_Gene_Lin <- paste0(data$Drug, "_", data$Gene, "_", data$Primary.Disease)
    print("Merged metadata.")
    # Merge mutation calls
    data <- unfactorize(data)
    data <- merge(data, maf_df, by = c("Gene", "Broad_ID"), all.x = TRUE)
    data$Mutation_Status <- ifelse(is.na(data$Mutation_Status), 0, data$Mutation_Status)
  }

  ## Summarize number of mutant and Wildtype cell lines
  data_muts_summ <- data %>% group_by(Gene) %>% summarize(N_Wildtype = length(Mutation_Status) - sum(Mutation_Status == 1), N_Mutant = sum(Mutation_Status == 1))
  ## Merge test results back into full dataset, which restores information lost in the summarization
  data <- merge(data_muts_summ, data, by = "Gene")
  print("Merged mutation calls.")

  # Merge GE data
  data <- merge(data, ge_df, by = c("Gene", "Broad_ID"), all.x = TRUE)
  print("Merged gene expression data.")

  # Merge CN data
  data <- merge(data, cn_df, by = c("Gene", "Broad_ID"), all.x = TRUE)
  print("Merged copy number data.")

  # Drop cell lines not in DepMap
  data <- filter(data, !is.na(Broad_ID) & !(Broad_ID %in% c("ACH-001182", "ACH-001451")))

  # Factorize gene
  data$Gene <- factor(data$Gene)

  # Save and return dataframe
  saveRDS(data, outfile, compress = "xz")
  return(data)
}

# FilterForPointMuts <- function(dataframe, outfile) {
#   # Filter for point mutations
#   ptmuts <- filter(dataframe, PointMutation == 1)
#   # Remove columns that lead to duplicate mutation entries
#   ptmuts <- subset(ptmuts, select = -c(Variant_Classification, Reference_Allele, Tumor_Seq_Allele1, PointMutation))
#   ptmuts$Gene <- ordered(ptmuts$Gene, levels = g2p_genes)
#   ptmuts <- unique(ptmuts)
#   saveRDS(ptmuts, outfile, compress = "xz")
#   return(ptmuts)
# }
```

@@

ccle <- MergeOmicsG2P(database = "CCLE", maf_df = maf_df, ge_df = ge_melt, cn_df = cn_melt, outfile = "./data_munging/rds/ccle_data_19Q1.rds")
# ctrp <- MergeOmicsG2P(database = "CTRP", maf_df = maf_df, ge_df = ge_melt, cn_df = cn_melt, outfile = "./../htscreens_giant_files/ctrp_data_19Q1.rds")
# gdsc <- MergeOmicsG2P(database = "GDSC", maf_df = maf_df, ge_df = ge_melt, cn_df = cn_melt, outfile = "./data_munging/rds/gdsc_data_19Q1.rds")

>>>>

---
title: "19Q2"
author: "Sally Claridge"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    theme: cerulean
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
    code_folding: hide
    pandoc_args: [
      "+RTS", "-K2048m",
      "-RTS"
    ]
---

@@

@@

This comprehensive cancer cell line information was curated by both Daniel Charytonowicz and myself.
```{r dan_ccl_database}
ccl_converter <- read.delim("./data_munging/cell_line_database_v1_20180911.tsv", row.names = 1, sep = "\t", header = TRUE)
ccl_lineage_converter <- unique(ccl_converter[, c("DepMap_ID", "grouped_general_doid", "group_general_lineage_name")])
lineage_converter <- unique(select(ccl_lineage_converter, grouped_general_doid, group_general_lineage_name))
ccl_histologies <- read.delim("./data_munging/ccl_histologies_metadata.csv", sep = ",", header = TRUE, row.names = 1)
```

@@

DepMap 19Q2 update fixed almost all of the inconsistencies of the Primary.Disease column, though there were come capitalization inconsistencies.
```{r depmap_metadata}
ccl_meta <- read.delim("./data_munging/19Q2/sample_info.csv", sep = ",", header = TRUE, na.strings = c("", NA))
ccl_meta <- select(ccl_meta, DepMap_ID, stripped_cell_line_name, CCLE_name, disease)
colnames(ccl_meta) <- c("DepMap_ID", "Cell_line", "CCLE_name", "disease")
# Remove two rows that contain cell lines with two different histologies
# The histologies that are kept are verified through matching in Cellosaurus and Cell Model Passports
# KMH2, haematopoietic_and_lymphoid: https://web.expasy.org/cellosaurus/CVCL_1330, https://cellmodelpassports.sanger.ac.uk/passports/SIDM01018
# MS1, lung: https://web.expasy.org/cellosaurus/CVCL_1429, https://cellmodelpassports.sanger.ac.uk/passports/SIDM00245
ccl_meta <- subset(ccl_meta, CCLE_name != "KMH2_THYROID" & CCLE_name != "MS1_SKIN")
ccl_meta <- subset(ccl_meta, !grepl("MERGED", CCLE_name, fixed = TRUE))
ccl_diseases <- as.character(unique(ccl_meta$disease))[as.character(unique(ccl_meta$disease)) != " "]
CCLE_histology <- tolower(gsub("^[a-z0-9]*_", "", ccl_meta$CCLE_name, ignore.case = TRUE))
ccl_meta <- add_column(ccl_meta, CCLE_histology = CCLE_histology, .after = "CCLE_name")
ccl_meta$disease <- with(ccl_meta, ifelse(disease == " " & Histology %in% ccl_diseases, as.character(Histology), as.character(disease)))
ccl_meta$disease <- with(ccl_meta, ifelse(disease == "ovary", "ovarian",
                                   ifelse(disease == " " & Histology == "vulva", "vulva",
                                   ifelse(disease == " " & Histology == "stomach", "gastric",
                                   ifelse(disease == " " & Histology == "upper_aerodigestive_tract", "upper_aerodigestive_tract",
                                   ifelse(disease == " " & Histology == "large_intestine", "colorectal",
                                   ifelse(disease == " " & Histology == "biliary_tract", "bile_duct",
                                   ifelse(disease == " " & Histology == "oesophagus", "esophagus",
                                   ifelse(disease == " " & Histology == "testis", "embryo", disease)))))))))
```

@@

```{r g2p}
g2p <- read.delim("./data_munging/g2p_db_edited.csv", sep = ",", header = TRUE, row.names = 1)
g2p$Drug_Gene_Lin <- gsub(pattern = " cancer", "", g2p$Drug_Gene_Lin)

@@

@@

Mutation calls (coding region, germline filtered):
```{r mutation_status, eval=FALSE}
maf_raw <- read.delim("./data_munging/19Q2/CCLE_mutations_20190514.csv.gz", header = TRUE, sep = ",")
maf_raw[, 1:2] <- NULL
colnames(maf_raw)[colnames(maf_raw) == "Hugo_Symbol"] <- "Gene"
maf_raw <- unfactorize(maf_raw)
# Select columns
maf <- subset(maf_raw, select = c("Gene", "DepMap_ID", "Variant_Classification", "Variant_Type"))
maf <- maf[!is.na(maf$Variant_Classification),]
maf <- maf[maf$Variant_Classification %in% unique(g2p$biomarker_type_curated),]
maf$Mutation <- 1
maf_df <- dcast(maf, Gene + DepMap_ID ~ Variant_Classification, fun.aggregate = mean, fill = 0)
maf_df_summ <- colSums(maf_df[, 3:ncol(maf_df)])

saveRDS(maf_df, "./data_munging/19Q2/maf_df_19Q2.rds", compress = "xz")
```

@@

```{r fusion_calls, eval=TRUE}
fusions <- read.delim("./data_munging/19Q2/CCLE_fusions.csv.gz", header = TRUE, sep = ",")
fusions$LeftGene <- gsub(" .*", "", fusions$LeftGene)
fusions$RightGene <- gsub(" .*", "", fusions$RightGene)
fusions$X.FusionName <- gsub("--", "-", fusions$X.FusionName)
fusions <- select(fusions, DepMap_ID, X.FusionName)
fusions <- as.data.frame(lapply(fusions, function(x) gsub("IGH@|IGH-@", "IGH", x)))
colnames(fusions) <- c("DepMap_ID", "FusionPair")
fusions <- merge(fusions, ccl_meta, by = "DepMap_ID", all.x = TRUE)
```

DepMap WES CN Data:
```{r copy_number, eval=FALSE}
cn <- read.delim("./data_munging/19Q2/CCLE_gene_cn_20190514.csv.gz", sep = ",", check.names = FALSE, header = TRUE)
# Remove Entrez gene IDs from colnames, format is "Gene (Entrez ID)"
colnames(cn) <- gsub(" .*", "", colnames(cn))
colnames(cn)[1] <- "DepMap_ID"
# Melt
cn_melt <- melt(data = cn, id.vars = "DepMap_ID", measure.vars = colnames(cn[2:ncol(cn)]), variable.name = "Gene", value.name = "CN")
cn_melt <- unfactorize(cn_melt)

saveRDS(cn_melt, "./data_munging/19Q2/cn_melt_19Q2.rds", compress = "xz")
```

19Q2, CCLE RNAseq gene expression data:
```{r gene_expression, eval=FALSE}
ge <- read.delim("./../htscreens_giant_files/CCLE_depMap_19Q2_TPM_20190305.csv.gz", header = TRUE, sep = ",", check.names = FALSE)
# Remove Entrez gene IDs from colnames, format is "Gene (Entrez ID)"
colnames(ge) <- gsub(" .*", "", colnames(ge))
colnames(ge)[1] <- "DepMap_ID"
# Melt
ge_melt <- melt(ge, id.vars = "DepMap_ID", measure.vars = colnames(ge)[2:ncol(ge)], variable.name = "Gene", value.name = "TPM")
ge_melt$TPM <- log2(ge_melt$TPM + 1)
ge_melt <- unfactorize(ge_melt)

saveRDS(ge_melt, "./data_munging/19Q2/ge_melt_19Q2.rds", compress = "xz")
```

Load Rds:
```{r load_omics_rds}
maf_df <- readRDS("./data_munging/19Q2/maf_df_19Q2.rds")
cn_melt <- readRDS("./data_munging/19Q2/cn_melt_19Q2.rds")
ge_melt <- readRDS("./data_munging/19Q2/ge_melt_19Q2.rds")
```

@@

```{r omics_functions}
MergeOmicsG2P <- function(database, maf_df, ge_df, cn_df, outfile) {
  # Merge mutation status, CN, and GE omics data with drugscreen results and metadata
  if(database %in% c("CCLE", "CTRP", "GDSC")) {
    filt <- auc %>% filter(Database == database) %>% select(Cell_line, Histology, TCGA_classification, CCLE_Name, DepMap_ID, Drug, AUC_IC50)
    print(paste0("Processing ", database, " data."))
    colnames(filt)[colnames(filt) == "AUC_IC50"] <- "Score"

    # Get all G2P gene-cell line combinations
    grid <- expand.grid("Cell_line" = unique(filt$Cell_line), "Gene" = g2p_genes)
    data <-  merge(filt, grid, by = "Cell_line", all.y = TRUE)
    print("Merged grid.")
    # Merge metadata
    data <-  merge(data, select(ccl_meta, Cell_line, Histology, CCLE_Name, DepMap_ID, Primary.Disease), by = c("Cell_line", "Histology", "CCLE_Name", "DepMap_ID"), all.y = TRUE)
    ## Add columns for G2P drug-gene and drug-gene-lineage sorting
    data$Drug_Gene <- paste0(data$Drug, "_", data$Gene)
    data$G2P_Drug_Gene <- ifelse(data$Drug_Gene %in% g2p_druggenes, "Yes", "No")
    data$Drug_Gene_Lin <- paste0(data$Drug, "_", data$Gene, "_", data$Primary.Disease)
    data$G2P_Drug_Gene_Lin <- ifelse(data$Drug_Gene_Lin %in% g2p_dgls, "Yes", "No")
    print("Merged metadata.")
    # Merge mutation calls
    data <- unfactorize(data)
    data <- merge(data, maf_df, by = c("Gene", "DepMap_ID"), all.x = TRUE)
    data$Mutation_Status <- ifelse(is.na(data$Mutation_Status), 0, data$Mutation_Status)
  }
  if(database == "CRISPR") {
    print(paste0("Processing ", database, " data."))
    # Melt CRISPR dataset for merging
    data <- melt(crispr_all, id.vars = "DepMap_ID", measure.vars = colnames(crispr_all)[2:ncol(crispr_all)], variable.name = "Gene", value.name = "Score")
    # Get all G2P gene-cell line combinations
    data <-  filter(data, Gene %in% g2p_genes)
    # Merge cell line metadata
    data <- merge(data, select(ccl_meta, Cell_line, Histology, CCLE_Name, DepMap_ID, Primary.Disease), by = "DepMap_ID", all.y = TRUE)
    data$Drug_Gene <- paste0(data$Drug, "_", data$Gene)
    data$Drug_Gene_Lin <- paste0(data$Drug, "_", data$Gene, "_", data$Primary.Disease)
    print("Merged metadata.")
    # Merge mutation calls
    data <- unfactorize(data)
    data <- merge(data, maf_df, by = c("Gene", "DepMap_ID"), all.x = TRUE)
    data$Mutation_Status <- ifelse(is.na(data$Mutation_Status), 0, data$Mutation_Status)
  }

  ## Summarize number of mutant and Wildtype cell lines
  data_muts_summ <- data %>% group_by(Gene) %>% summarize(N_Wildtype = length(Mutation_Status) - sum(Mutation_Status == 1), N_Mutant = sum(Mutation_Status == 1))
  ## Merge test results back into full dataset, which restores information lost in the summarization
  data <- merge(data_muts_summ, data, by = "Gene")
  print("Merged mutation calls.")
  # Merge GE data
  data <- merge(data, ge_df, by = c("Gene", "DepMap_ID"), all.x = TRUE)
  print("Merged gene expression data.")
  # Merge CN data
  data <- merge(data, cn_df, by = c("Gene", "DepMap_ID"), all.x = TRUE)
  print("Merged copy number data.")
  # Drop cell lines not in DepMap
  data <- filter(data, !is.na(DepMap_ID) & !(DepMap_ID %in% c("ACH-001182", "ACH-001451")))
  # Save and return dataframe
  data <- unfactorize(data)
  saveRDS(data, outfile, compress = "xz")
  return(data)
}
```

@@

ccle <- MergeOmicsG2P(database = "CCLE", maf_df = maf_df, ge_df = ge_melt, cn_df = cn_melt, outfile = "./data_munging/rds/ccle_data_19Q2.rds")
# ctrp <- MergeOmicsG2P(database = "CTRP", maf_df = maf_df, ge_df = ge_melt, cn_df = cn_melt, outfile = "./../htscreens_giant_files/ctrp_data_19Q2.rds")
# gdsc <- MergeOmicsG2P(database = "GDSC", maf_df = maf_df, ge_df = ge_melt, cn_df = cn_melt, outfile = "./data_munging/rds/gdsc_data_19Q2.rds")
```
